---
title: "Seaflow Analysis"
output: html_notebook
---

# Seaflow Analysis of Phytoplankton Carbon Content Using Flow Cytometry

# Introduction

Flow cytometry allows researchers to measure physical and chemical properties of cells as they pass through a laser. In this analysis, we aim to quantify carbon content per cell for different phytoplankton species using forward scatter data, which reflects cell size and shape. This study compares two flow cytometry instruments, 740 and 751, to evaluate if these measurements can help predict carbon content, a crucial factor in understanding the role of phytoplankton in carbon cycling.

# Setup and Library Loading

These libraries help streamline the analysis, manipulation, and visualization of data, making the code cleaner and easier to manage.

```{r}
library(tidyverse)    # For data manipulation}
library(scales)       # For color transparency
library(viridis)      # For color palette
library(gridExtra)    # To arrange multiple plots in a single output
library(purrr)
library(dplyr)        #for data manipulation, especially with lists and pipes
```

# Setting the working directory

This function sets the working directory, which is the folder where your R session will look for files and save outputs. By setting this path, you ensure that all file reading and writing operations are performed relative to this directory. This is particularly useful when you are working in a project folder, ensuring all files (like datasets and outputs) are organized in one place. Adjust this path as needed for your setup.

```{r}
path.to.git.repository <- "~/Documents/Codes/fsc-poc-calibration"
setwd(path.to.git.repository)
```

# Loading the data

```{r}
poc <- read.csv("Qc-cultures.csv")
```

# Loading and Processing Culture Data

The first block of code loops through two instruments (740 and 751), loading culture data specific to each instrument from CSV files named according to the instrument number (e.g., 740-cultures.csv for instrument 740).

```{r}
# Define a vector of instrument IDs (740 and 751) to load datasets for each instrument.
instruments <- c("740", "751")

# Loop through each instrument in the vector 'instruments'.
for (inst in instruments) {
  # Dynamically create the file name for each instrument and read in the corresponding data.
  # Suppress column type output to keep the notebook clean.
  cultures <- read_csv(paste0(inst, "-cultures.csv"), show_col_types = FALSE)
  
  # Print the data specification for 'cultures' to check the column types of each dataset.
  print(spec(cultures))
}

```

Each instrument generates data with different characteristics (e.g., fluorescence, forward scatter, side scatter) which we want to analyze and compare. The read_csv() is used to load the data and print(spec(cultures)) is used to check the data types, ensuring that numeric and categorical values are appropriately recognized.

# Assigning Sample IDs Based on Instrument

Next, we assign specific sample identifiers (Sample.ID) for each instrument to ensure consistency in labeling.

```{r}
cultures <- cultures %>%
  mutate(Sample.ID = case_when(
    inst == 740 ~ rep(c("Thalassiosira weissflogii", "Navicula transitans", 
                        "Thalassiosira pseudonana (1135)", "Thalassiosira pseudonana (3367)", 
                        "Phaedactylum tricornutum", "Micromonas pusilla", 
                        "Prochlorococcus (AS9601)", "Prochlorococcus (1314)", 
                        "Prochlorococcus (MED4)", "Prochlorococcus (NATL12A)", 
                        "Synechococcus (WH8102)", "Synechococcus (WH7803)"), each = 2),
    inst == 751 ~ rep(c("Thalassiosira weissflogii", "Navicula transitans", 
                        "Thalassiosira pseudonana (1135)", "Thalassiosira pseudonana (3367)", 
                        "Phaedactylum tricornutum", "Micromonas pusilla", 
                        "Prochlorococcus (MED4)", "Prochlorococcus (AS9601)", 
                        "Prochlorococcus (1314)", "Prochlorococcus (NATL12A)", 
                        "Synechococcus (WH8102)", "Synechococcus (WH7803)"), each = 2)
  ))

```

# Calculating Standard Deviation and Mean for Numeric Data

For each sample, we calculate the standard deviation and mean of the numeric data (e.g., forward scatter, chlorophyll content, carbon content).

```{r}
# Calculate the standard deviation for each numeric group in 'cultures'
cultures_sd <- cultures %>%
  group_by(Sample.ID) %>%
  summarize(across(where(is.numeric), ~ sd(.x, na.rm = TRUE)), .groups = 'drop') %>%
  rename_with(~ paste0(.x, ".sd"), -Sample.ID)

# Calculate the mean for each numeric group in 'cultures'
cultures_mean <- cultures %>%
  group_by(Sample.ID) %>%
  summarize(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)), .groups = 'drop')

# Add the standard deviation values as new columns in 'cultures_mean'
  cultures_mean <- cultures_mean %>%
    inner_join(cultures_sd, by = "Sample.ID") %>%
    rename(norm.fsc.sd = norm.fsc.sd, norm.chl.sd = norm.chl.sd)

```

The standard deviation (sd) measures the variability in the measurements, and the mean (mean) represents the central tendency of the data. These calculations are critical for understanding how consistent or variable the measurements are across samples and for error propagation in future analyses.

# Merging with POC Data

Merging Culture Data with POC (Particulate Organic Carbon) Data The poc dataset contains information on particulate organic carbon, which we want to link to the flow cytometry data by matching Sample.ID. This combined dataset allows for a comprehensive analysis of flow cytometry measurements (such as forward scatter) and biological metrics (such as carbon content per cell).

```{r}
# Merge with 'poc' data on 'Sample.ID' to combine related measurements
merge_data <- poc %>%
  inner_join(cultures_mean, by = "Sample.ID")

 # Write the final merged data to a CSV file for the current instrument
  merge_data %>%
    select(Sample.ID, norm.fsc, norm.fsc.sd, norm.chl, norm.chl.sd, 
           abundance_cells_mL_mean, abundance_cells_mL_sd, pgC_cell, pgN_cell, 
           pgC_cell_sd, pgN_cell_sd) %>%
    write_csv(paste0(inst, "-Qc-cultures.csv"))
}
```

Merging ensures that we have a unified dataset that includes both flow cytometry measurements (e.g., forward scatter, chlorophyll) and the biological measurements (e.g., cell abundance, carbon content) for the same sample. This allows for more comprehensive analyses, such as understanding the relationship between flow cytometry metrics and biological content.

# Loading Mie data

Ensure that the file is present in your working directory.

```{r}
# Load the Mie theory data from CSV file
mie <- read_csv("calibrated-mie.csv")
```

# Creating and Saving Plots

Scatter plots show the relationship between normalized forward scatter (size indicator) and carbon content per cell. Error bars highlight variability, and a red line represents the Mie scattering model, a theoretical model for predicting cell carbon content based on scatter.

```{r}
# Define output PNG for side-by-side plots, double the width
png("Qc-scatter.png", width = 18, height = 8, units = "in", res = 300)

# Initialize list to store plots
plots <- list()

# Loop through each instrument (740 and 751)
for (inst in c(740, 751)) {
  
  # Load culture data for each instrument
  merge <- read_csv(paste0(inst, "-Qc-cultures.csv"))
  
  merge2 <- merge %>%
    dplyr::filter(Sample.ID != "Phaeodactylum tricornutum") %>%
    arrange(norm.fsc)
  
  # Create a ggplot for the current instrument
  p <- ggplot(merge2, aes(x = norm.fsc, y = pgC_cell)) +
    
    # Add error bars for pgC.cell with standard deviations
    geom_errorbar(aes(ymin = pgC_cell - pgC_cell_sd, ymax = pgC_cell + pgC_cell_sd), 
                  width = 0, color = "grey", size = 1) +
    geom_errorbarh(aes(xmin = norm.fsc - norm.fsc.sd, xmax = norm.fsc + norm.fsc.sd),
                   height = 0, color = "grey", size = 1) +
    
    # Add lines for Mie-based model with upper, mid, and lower bounds
    geom_line(data = mie, aes(x = scatter, y = !!sym(paste0("Qc_", inst, "_mid")), 
                              color = "Mie-based model (n = 1.38 ± 0.3)"), 
              size = 1.2) +
    geom_line(data = mie, aes(x = scatter, y = !!sym(paste0("Qc_", inst, "_upr"))), 
              color = "grey", size = 1, show.legend = FALSE) +
    geom_line(data = mie, aes(x = scatter, y = !!sym(paste0("Qc_", inst, "_lwr"))), 
              color = "grey", size = 1, show.legend = FALSE) +
    
    # Add larger points for each culture sample with semi-transparent colors
    geom_point(aes(fill = Sample.ID), shape = 21, size = 8, 
               color = "black", alpha = 0.7) +  # Larger points and slightly more opaque
    scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.9) +
    
    # Custom legend entry for the Mie-based model
    scale_color_manual(values = c("Mie-based model (n = 1.38 ± 0.3)" = "red3")) +
    
    # Set log scales for x and y axes
    scale_x_log10(limits = c(0.002, 10), 
                  breaks = c(0.002, 0.005, 0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10),
                  labels = scales::label_number()) +
    scale_y_log10(limits = c(0.005, 100),
                  breaks = c(0.005, 0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100),
                  labels = scales::label_number()) +
    
    # Add labels and title
    labs(x = "Normalized scatter (dimensionless)", 
         y = expression(paste("Qc (pgC cell"^{-1},")")),
         title = paste("Instrument:", inst)) +
    
    # Customize theme for clarity and format
    theme_classic(base_size = 14) +
    theme(
      panel.border = element_rect(color = "black", fill = NA, size = 1),
      legend.position = c(0.2, 0.7),  # Position legend in the top-left corner
      legend.background = element_blank(),  # No legend background box
      legend.title = element_blank(),
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12)
    )
  
  # Save the plot in the list
  plots[[as.character(inst)]] <- p
}
```

# Arranging and Saving Multiple Plots

Finally, the plots for both instruments (740 and 751) are arranged side by side in a single PNG file for comparison.

```{r}
grid.arrange(grobs = plots, ncol = 2)
dev.off()
```

# Linear regression

This code focuses on analyzing flow cytometry data to explore correlations between normalized forward scatter values (norm.fsc) and carbon content per cell (pgC_cell). The code loops through each instrument to calculate correlation and regression statistics, visualizes the data with scatter plots, using error bars and reference model lines, performs regression analysis on both original and log-transformed data.

```{r}

# Initialize list to store plots
plots <- list()

generate_plot <- function(inst, merge2, mie) {
  # Generate column name dynamically
  col_name <- paste0("Qc_", inst, "_mid")
  
  # Match scatter intervals to mie scatter values
  id <- findInterval(merge2$norm.fsc, mie$scatter)
  id[is.na(id)] <- 1  # Replace NA indices with a valid default
  
  # Add predicted values to merge2 using the mie table
  merge2 <- merge2 %>%
    mutate(predicted = mie[id, col_name] %>% pull())  # Ensure numeric type
  
  # Linear regression
  reg <- lm(pgC_cell ~ predicted, data = merge2)
  print(paste("Regression summary for instrument", inst))
  print(summary(reg))
  

```

This function (generate_plot) automates the process of creating scatter plots for each instrument, using dynamic column naming to allow for specific regression modeling. Data Matching: The code matches the norm.fsc values in merge2 to Mie model values in mie to get predicted Qc values. Linear Regression: A linear regression of pgC_cell against predicted Qc values is used to assess how well the predicted values align with measured data.

# Plotting Data with Scatter Plots and Error Bars

```{r}
  # Generate the ggplot
  p <- merge2 %>%
    ggplot(aes(x = norm.fsc, y = pgC_cell)) +
    
    # Add vertical error bars
    geom_errorbar(aes(ymin = pgC_cell - pgC_cell_sd, ymax = pgC_cell + pgC_cell_sd), 
                  color = "grey", width = 0) +
    
    # Add horizontal error bars
    geom_errorbarh(aes(xmin = norm.fsc - norm.fsc.sd, xmax = norm.fsc + norm.fsc.sd), 
                   color = "grey", height = 0) +
    
    # Add larger points with transparency
    geom_point(aes(fill = Sample.ID), size = 8, shape = 21, alpha = 0.7) +
    
    # Set color scale for points with title for the legend
    scale_fill_viridis_d(name = "Sample ID") +
    
    # Log scale for x and y axes with limits and breaks
    scale_x_log10(limits = c(0.002, 10), 
                  breaks = c(0.002, 0.005, 0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10)) +
    scale_y_log10(limits = c(0.005, 100), 
                  breaks = c(0.005, 0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50, 100)) +
    
    # Axis labels and plot title
    labs(x = "Normalized scatter (dimensionless)", 
         y = expression(paste("Qc (pgC cell"^{-1}, ")")),
         title = paste("Instrument", inst)) +
    
    # Add red reference line for the Mie-based model
    geom_abline(intercept = 0.5239, slope = 1.0462, color = "red3", size = 1) +
    
    # Apply classic theme to add a border around the plot
    theme_classic(base_size = 14) +
    
    # Customize theme elements for legend, plot border, and text size
    theme(
      panel.border = element_rect(color = "black", fill = NA, size = 1),  # Border around the plot
      legend.position = c(0.2,0.7),                                          # Position legend on the right
      legend.background = element_blank(),                                # No background box for the legend
      legend.title = element_text(size = 16),                             # Increase legend title font size
      legend.text = element_text(size = 14),                              # Increase legend text font size
      plot.title = element_text(hjust = 0.5, size = 16),                  # Centered and larger title
      axis.text = element_text(size = 12),                                # Axis text size
      axis.title = element_text(size = 14),                               # Axis title size
      panel.grid.major = element_line(color = "grey", size = 0.5),        # Add major grid lines
      panel.grid.minor = element_line(color = "lightgrey", size = 0.25)   # Add minor grid lines
    ) +
    
    # Adjust legend to control point size in the legend
    guides(fill = guide_legend(title = "Sample ID", override.aes = list(size = 5)))
  
  # Return the plot object
  return(p)
}

# Generate plots for both instruments
plots <- map(c("740", "751"), ~generate_plot(.x, merge2, mie))

# Save all plots in a single PNG file using gridExtra
png("Qc-scatter-combined.png", width = 24, height = 10, units = "in", res = 300)
grid.arrange(grobs = plots, ncol = 2)  # Arrange plots side by side
dev.off()

```

# Regression Analysis on Log-Transformed Data

The relationship between forward scatter and carbon content might be better explained on a log-log scale, particularly if the data spans several orders of magnitude. This section applies log transformation to both variables, performs a linear regression, and then prints the summary, providing insight into how well forward scatter can predict carbon content.

```{r}
# Log-transformed regression for both instruments
for (inst in instruments) {
  log_reg <- merge2 %>%
    dplyr::filter(pgC_cell > 0, norm.fsc > 0) %>%  # Remove invalid values for log
    mutate(log_pgC.cell = log10(pgC_cell), log_norm.fsc = log10(norm.fsc)) %>%
    lm(log_pgC.cell ~ log_norm.fsc, data = .)
  
  print(paste("Log-transformed regression summary for instrument", inst))
  print(summary(log_reg))
}

```

Data Preparation for Log Transformation: Invalid values (pgC_cell or norm.fsc \<= 0) are filtered out to ensure log transformation can be performed without issues. Log-Transformed Regression: A linear regression is conducted on the log-transformed data, allowing for interpretation of the relationship between norm.fsc and pgC_cell in a log-log scale.
